# Rstudio exercise 6

The task in the Rstudio exercise 6 is to repeat the analyses done in chapters 8 and 9 of MABS (Vehkalahti and Everitt, 2019), but use the two datasets in reverse order. In chapter 8 of MABS, the analyses are done with BPRS data, but in this exercise we implepent the analyses using RATS data. Similarly, we implement the analyses of chapter 9 of MABS using the BPRS data even though they were originally done with RATS data. To make my report easier to follow, I have used same numbers for figures and tables as were used in th chapters of MABS.

## The analyses of Chapter 8 of MABS using RATS data

``` {r, warning=FALSE, message=FALSE}
# accessing the data and re-converting ID and Group as factors

RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep = '\t', header = T)

RATS$ID <- factor(RATS$ID)
RATS$Group <- factor(RATS$Group)

RATSL <- read.csv(file="~/GitHub/IODS-project/data/RATSL.csv")

RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
```


``` {r, message=FALSE, warning= FALSE}
library(ggplot2)

# Drawing the figure 8.1 using RATS data
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```

##### **FIGURE 8.1** Individual growth by diet (Group) for the RATS data

In the Figure 8.1 all weights from all time points are plotted separately for each diet group. Each line represents one rat. It can easily be seen, that the baseline weight in Group 1 is a lot lower compared to groups 2 and 3. The rats in group 1 also don's seem to grow much. The lines in each group seem very aligned, so there is not much invidiual variablitiy.

In next plot (Figure 8.2) the standardized values of each observation are plotted. The standardiced values were calculated by subtracting the relevant occasion mean from the original observation and then dividing by the corresponding visit standard deviation. From the scaled plot, the effect of the baseline value is more easily seen, and now we see some individual variablity more clearly. For example, the solid line in Group 2 has quite high weight at baseline and a bit slower groth compared to other rats in that group. In the plot using standardized values, the standardized weight recreases rather than increases.

```{r, warning=FALSE, message = FALSE}
library(dplyr)

# Standardising the weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight) ) %>%
  ungroup()

# Plotting again with the standardised weight
ggplot(RATSL, aes(x = Time, y = stdWeight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(name = "standardized weight")
```

##### **FIGURE 8.2** Individual growth by diet for RATS data after standardization.

```{r, message=FALSE, warning=FALSE}
library(tidyr)

# Number of timepoints, baseline included
n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
RATSL_summary <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Plot the mean profiles
ggplot(RATSL_summary, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")
```

##### **FIGURE 8.3** Mean growth for the diets in the RATS data.

In the Figure 8.3 the mean growth for the three diet groups is presented. Same can be seen here as earlier, the mean weight of rats in Group 1 is lower, as well as the mean growth rate seems to be lower. Other approach to plot summaries is to use boxplots (Figure 8.4). 

```{r, message=FALSE, warning=FALSE}
#Drawing the figure 8.4 using RATS data

ggplot(RATSL, aes(x = factor(Time), y = Weight, fill = Group))+ 
  geom_boxplot(position = position_dodge(width = 0.9)) +
  theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = c(0.9,0.4)) +
  scale_x_discrete(name = "Time") +
  scale_y_continuous(name = "Weight") +
  scale_fill_grey(start = 0.5, end = 1)
```

##### **FIGURE 8.4** Boxplots for summaries of weight for the diets at each time point in the RATS data.


Boxplots gives also a possiblity to identify outliers. In group 2 there is one rat with considerable higher weight than others which can bee seen from the figure 8.5 where mean weights of each rat at all timepoints after baseline are plotted. I'm quite hesitant to identify this rat as an otlier, because the weight is clearly at possible range and from the earlier figures we see that the growth of this heavy rat is similar compared to other is that group. However, for educational purposes, I filter out this rat and draw the figure 8.6 without it.

```{r, message=FALSE, warning=FALSE}
#Drawing the figure 8.5 using RATS data

RATSL_summary2<- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise(mean=mean(Weight) ) %>%
  ungroup()

ggplot(RATSL_summary2, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), all time points after baseline")
```

##### **FIGURE 8.5** Boxplots of mean weight for the diet groups in the RATS data.

```{r, message=FALSE, warning=FALSE}
# Filtering out the heavy rat in group 2 and drawing the figure 8.6
RATSL_summary2F <- RATSL_summary2 %>% filter (mean <560)

ggplot(RATSL_summary2F, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), all time points after baseline")
```

##### **FIGURE 8.6** Boxplots of mean weight for the diet groups in the RATS data without an outlier (a heavy rat in Group 2).


The differences between diet groups can't be tested with t-test as was done in chapter 8 for BPRS data, because t-test can only deal with two groups. The analysis of covariance (Table 8.4) is possible. From the ANCOVA table (Table 8.4) we see that the mean weights in diet groups are not different when the baseline weight is taken into account (p=0.076).

##### **TABLE 8.4** Results from an analysis of covariance of the RATS Data with Baseline Weight and diet group covariates
```{r, message=FALSE, warning=FALSE}
# Add the baseline weight from the original data as a new variable to the summary data:
RATSL_summary2b <- RATSL_summary2 %>%
  mutate(baseline = RATS$WD1)
# Fit the ANCOVA model and see the results:
fit <- lm(mean ~ baseline + Group, data = RATSL_summary2b)
anova(fit)
```


## The analyses of Chapter 9 of MABS using BPRS data

First, we explore the data ignoring the repeated-measures structure (Figure 9.1 and Table 9.3).
``` {r, warning=FALSE, message=FALSE}
# accessing the data and re-converting treatment and subject as factors

BPRS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt", sep  =" ", header = T)

BPRS$treatment <- factor(BPRS$treatment)
BPRS$subject <- factor(BPRS$subject)

BPRSL <- read.csv(file="~/GitHub/IODS-project/data/BPRSL.csv")

BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
```

```{r, warning=FALSE, message=FALSE}
# Drawing the Fig 9.1
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_text(aes(label = treatment)) +
  scale_x_continuous(name = "Week", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs score") + theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

##### **FIGURE 9.1** Bprs score plotted agains weeks ignoring the repeated measures structure. The treatment to which each observations belongs identified with number.

Results from linear regression ignoring the repeated-measures structure (Table 9.3) show that treatment is not significantly associated with bprs score (p=0.661), but the regression on week is higly significant (p<0.001).

##### **TABLE 9.3** Linear regression model to BPRS data in *long form* with BPRS scores as target variable and treatment and week as explanatory variables. Ignoring the repeated-measures structure of the data.
```{r, message = FALSE, warning=FALSE}
BPRS_reg <- lm(bprs ~ treatment + week, data = BPRSL)
summary(BPRS_reg)
```


The figures taking the repeated-measures structure into account are presented below (Figure 9.2 and Figure 9.3). From the figure 9.2 we see the individual changes in bprs scores. In the original figure 9.2, all groups (in this case treatments) were plotted in the same picture. I divided them into two figures according to treatment because otherwise figure would have been too messy. From the scatterplot matrix (Figure 9.3) we can see that repeated measures are dependent from one another.

```{r, warning=FALSE, message=FALSE, fig.width=10, fig.height=5}
# Drawing the Fig 9.2
ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
  geom_line(aes(linetype = treatment)) +
  scale_x_continuous(name = "Time (weeks)", breaks = seq(0, 8, 1)) +
  scale_y_continuous(name = "bprs score") +
  theme_bw() + theme(legend.position = "none") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  facet_grid(. ~ treatment, labeller = label_both)
```

##### **FIGURE 9.2** The cange in bprs scores in individual participants.

```{r, fig.width=10, fig.height=10}
#Drawing the Fig 9.3
pairs(BPRS[, 3:11], cex = 0.7)
```

##### **FIGURE 9.3** Scatterplot matrix of repeated measures in BPRS data.

Next we proceed fitting linear mixed models to the BPRS data. First step is to fit a random interept model with week and treatment as explanatory variables (Table 9.4). Random intercept model means that we allow the intercepts of the fitted regression vary between subjects. From the results we see that variance of intercepts is 47.41 whereas residual variance is 104.21. Even though the residal variance is higher than the variance of intercepts, the variance of intercepts is still considerable. And also based on graphical examination of the data, it is reasonable to assume intercepts to vary between sucjects. The parameter estimates for week and treatment did not change much compared to the independence model (Table 9.3).

##### **TABLE 9.4** Results from fitting random intercept model, with week and treatment as explanatory variables, to BPRS data
```{r, warning=FALSE, message=FALSE}
#Table 9.4
library("lme4")
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref)
```

Second step is to fit a random intercept and slope model. Now we allow also the slope of the regression line vary between subjects. Parameter estimates for fixed effects (week and treatment) are again very similar as with earlier models. We can also see that the variance of slopes is not very high (0.69). However, allowing random slopes increased the variance of intercepts and also comparing the models with ANOVA indicated slightly  better fit for the second model (p=0.026).

##### **TABLE 9.5** Results from fitting random intercept and slope model, with week and treatment as explanatory variables, to BPRS data
```{r}
# Table 9.5
BPRS_ref2 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref2)

anova(BPRS_ref2, BPRS_ref)
```

The final step is to fit a random intercept and slope model allowing a treatmet + week interaction (Table 9.6). Now we can see some changes also in parameter estimates of fixed effect. However, the interaction between week and treatment does not appear very strong (beta 0.72 s.e. 0.40, T-value 1.785). Also when this final model is compared with previous model in ANOVA, it does not appear have significantly better fit to the data (p=0.075). The final conlusion from the models is that bprs scores decrease over time, but there is no difference between treatment groups.

##### **TABLE 9.6** Results from fitting the random intercept and slope model that allows for a treatment + week interaction to BPRS data

```{r}
# TAble 9.6
BPRS_ref3 <- lmer(bprs ~ week*treatment + (week | subject), data = BPRSL, REML = FALSE)
summary(BPRS_ref3)

anova(BPRS_ref3, BPRS_ref2)
```

Fitted cahnges in bprs scores from the final model and observed bprs changes are plotted in the figure 9.4. The figures support the conlusions made from the linear mixed models. The individual slopes show a clear decrease in bprs scores over time (weeks), but there seems to be no difference between the treatment groups 1 and 2. Steepness of the slopes seem to be approximately equal in both treatment groups.

```{r, fig.width=8, fig.height=4}
Fitted <- fitted(BPRS_ref3)
BPRSL <- BPRSL %>% mutate(Fitted)
graph1 <- ggplot(BPRSL, aes(x = week, y = bprs, group = subject)) +
 geom_line(aes(linetype = treatment)) +
 scale_x_continuous(name = "week", breaks = seq(0, 8, 1)) +
 scale_y_continuous(name = "bprs score", limits= c(15,95), breaks = seq(15, 95, 20)) +
 theme_bw() + theme(legend.position = "none")  +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
 ggtitle("Observed") +
  facet_grid(. ~ treatment, labeller = label_both)


graph2 <- ggplot(BPRSL, aes(x = week, y = Fitted, group = subject)) +
 geom_line(aes(linetype = treatment)) +
 scale_x_continuous(name = "week", breaks = seq(0, 8, 1)) +
 scale_y_continuous(name = "bprs score", limits= c(15,95), breaks = seq(15, 95, 20)) +
 theme_bw() + theme(legend.position = "none")  +
 theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
 ggtitle("Fitted") +
  facet_grid(. ~ treatment, labeller = label_both)

graph1; graph2
```

##### **FIGURE 9.4** Fitted changes in bprs scores from the interaction model and observed bprs changes


